---
title: "p8105_hw3_jl7035"
author: "Jeffrey Lin"
date: "2024-10-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Necessary Libraries 

```{r}
library(tidyverse)
library(ggplot2)
library(ggridges)
library(patchwork)
```
## Problem 1

Load Data
```{r}
library(p8105.datasets)
data("ny_noaa")

```

Inspect Data
```{r}
head(ny_noaa)

```
There are `r nrow(ny_noaa)` data points and `r ncol(ny_noaa)` variables of 
interest in this dataset. The key variables in this dataset are the 
id of the weather station, the date of which data was collected, amount 
of precipitation, amount of snow, depth of the snow, and the minimum and 
maximum temparature on that given date. Examining the number of missing values 
in this dataset, I would argue that they are a valid concern. With a large 
amount of missing values, any plots made may not accurately represent 
real life conditions. This conclusion is driven by the number of missing 
values in this dataset, which is `r sum(is.na(ny_noaa))`. Naturally, 
merely summing the amount of NA values it not necessarily the most accurate 
measure, as it doesn't capture how NA is spread across the columns.

Clean the Data
```{r}
ny_noaa <- ny_noaa %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
  ) %>%
  separate(date, into = c("year", "month", "day"), sep = "-")

head(ny_noaa)


```
Here I noticed that tmax and tmin were originally char types, which is why
I changed them to numeric. The separation of the date into 3 columns was 
per the instructions of the spec. 


Find most commonly observed values for snowfall 
```{r}
ny_noaa %>% 
  count(snow) %>% 
  arrange(desc(n))
  
```
The most commonly occurring value for snowfall is 0. This likely because the 
wide majority of days per year do not see snowfall. Afterwards, the second
most frequently appearing value is NA, indicating missing data. Afterwards, 
we also see values like 25, 13, 51, and 76, however I am not sure why these 
values are so commonly occurring.

Make a two-panel plot showing the average max temperature in January and in July 
in each station across years. Is there any observable / interpretable structure? 
Any outliers?
```{r}
ny_noaa %>% 
  group_by(id, month, year) %>% 
  filter(month == '01'|month == '07') %>% 
  summarise(avg_max_temp = mean(tmax, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = avg_max_temp, group = id)) +
  geom_point() +
  facet_grid(~month)



```
Examining this plot, we see the general trend that across all the reporting 
weather stations, the average max temperature seen is higher in July than 
January. This trend remains consistent throughout the entire span of years. 
There are a couple of outlier cases. The most notable appears to occur in 
July of 1987, where there was an uncharacteristically low average max 
temperature reported by a weather station.


Make a two-panel plot showing (i) tmax vs tmin for the full dataset 
(note that a scatterplot may not be the best option); and 
(ii) make a plot showing the distribution of snowfall values 
greater than 0 and less than 100 separately by year

```{r}
tmax_vs_tmin <- ny_noaa %>% 
  ggplot(aes(x = tmax, y = tmin)) +
  geom_hex()

snowfall_less_100 <- ny_noaa %>%
  group_by(year) %>% 
  filter(snow > 0 & snow < 100) %>% 
  ggplot(aes(x = year, y = snow)) +
  ylab("snow fall")
  geom_density_ridges()
  
tmax_vs_tmin + snowfall_less_100

```




## Problem 2

Load in and clean each dataset
```{r}
demographic_df <- 
  read_csv("Data/nhanes_covar.csv",
    na = c("", "NA", "."),
    skip = 4
  ) %>% 
  janitor::clean_names() %>% 
  mutate(
    sex = factor(
      sex, 
      levels = c("1", "2"), 
      labels = c("male", "female")),
    education = factor(
      education,
      levels = c("1", "2", "3"),
      labels = c("Less than high school", "High School Equivalent", 
                 "More than high school")
    )
  ) %>% 
  filter(
    age >= 21
  ) %>% 
  drop_na()
  
head(demographic_df)
  
accelerometer_df <- 
  read_csv("Data/nhanes_accel.csv",
    na = c("NA", "", ".")
  ) %>%  
  janitor::clean_names() %>%
  pivot_longer(
    min1:min1440,
    names_to = "Minute",
    values_to = "MIMS"
  ) %>% 
  mutate(
    Minute = as.numeric(str_remove(Minute, "min"))
  )
  

```
Join dfs together
```{r}
accel_participant_df <-
  left_join(
    demographic_df, 
    accelerometer_df, 
    by = "seqn")

anti_join(demographic_df, accel_participant_df)
```


Produce reader-friendly table for number of men and women in each education
category
```{r}
demographic_df %>% 
  group_by(sex, education) %>% 
  count() %>% 
  pivot_wider(
    names_from = education,
    values_from = n
  ) %>% 
  knitr::kable()

```
Looking at the count of the number of men and women in each education category,
we can see that there only appears to be a major difference in the number of 
men and women high school equivalent category, such that there were 12 more 
male participants than female participants with a high school equivalent 
education. Interestingly, more female participants had more than high school
education compared to men, but only by a difference of +3.

create visualization of age distribution for each education category
```{r}
ggplot(demographic_df, aes(x = education, y = age, fill = sex)) +
  geom_boxplot(outliers = TRUE)

```
Examining the distribution of age by educational level for male and female 
participants. We can once again notice some differences. Firstly, we can note 
that while the median age of men and women in the less than high school and more
than high school categories are relatively similar, we see a more substantial 
difference in the high school equivalent category. Here, female participants 
have a median age hovering around 62-63, which is higher than the male 
equivalent, at roughly 54. Looking at the spread of values in each of the 3 
categories, we can notice, perhaps a tiny bit of skew. In the less than 
high school and high school equivalent categories, for both men and women,
the distance between median and q3 is slightly smaller and suggestive of a 
potential slight left skew. For the more than high school equivalent, the median
is closer to q1, and potentially suggests a slight right skew. Lastly, we can 
see that the whole distribution of ages for the female high school equivalent 
category is shifted upwards compared to the male counterpart, as the q1 and q3 
are higher in the female category vs the male. 

compare total activity for men and women and by education
```{r}
demographic_df <- accel_participant_df %>% 
  group_by(seqn) %>% 
  select(MIMS, seqn) %>% 
  summarise(total_activity = sum(MIMS)) %>% 
  left_join(demographic_df, by = "seqn") %>% 
  relocate(total_activity, .after = education)

ggplot(demographic_df, aes(x = age, y = total_activity, color = sex)) +
  xlab("Age of Participant") +
  ylab("Total MIMS") +
  geom_point( alpha = 0.5) +
  geom_smooth(aes(color = sex), se = FALSE) +
  facet_grid(.~education)


```
Looking at these 3 plots, firstly it must be noted that there seems to be great
variation in the total activity levels by age for both sexes, so the trend lines
do not fit with the data perfectly. Now, comparing the 3 categories, it is 
interesting to note that participants in the less than high school education 
category start with the highest Total MIMS at ages close to 20, but see a 
sharp decline by age 80, such that they seem to be lower than the other two 
educational categories. Now looking at each of the three educational categories,
for less than high school, it seems that female participants have more total 
activity when they are younger, but men surpass them around age and even see 
a resurgent increase in activity around 60. Ultimately, both sexes finish with 
similar activity levels in old age. For high school equivalent, female 
participants consistently show more activity than their male counterparts, 
until they reach age 80. Interestingly, women see upticks in activity around 
ages 40 (global maximum) and 70 (local maximum). Lastly, in the more than high
school education category, female participants show more activity than men 
consistently until age 80. Women seem to peak in total MIMS when nearing age
60, before waning afterwards. Men reach their peak activity earlier, around 45,
before decreasing afterwards, albeit at a less steep rate than their female 
counterparts.

Create df representing hourly activity of participants
```{r}
daily_activity <- accel_participant_df %>% 
  mutate(hour = ceiling(Minute / 60)) %>% 
  group_by(seqn, hour) %>%
  summarise(hourly_activity = sum(MIMS)) %>% 
  left_join(demographic_df) %>% 
  relocate(hourly_activity, .before = total_activity)
```

24-hour activity time courses for each education level, shown by sex
```{r}
ggplot(daily_activity, aes(x = hour, y = hourly_activity, color = sex)) +
  geom_point(alpha = 0.2) +
  geom_smooth(aes(color = sex), se =FALSE) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 4),
    limits = c(0,24)) +
  viridis::scale_color_viridis(
    discrete = TRUE
  ) +
  xlab("Hour of the Day") +
  ylab("Activity by Hour") +  
  facet_grid(.~education) 


```
Looking at the graphs, we notice a similar pattern of hourly activity across 
the 24 hours for all 3 educational categories and for both sexes. Activity 
tends to be the lowest in the early hours of the day, hitting a global 
minimum around hour 4. Afterwards, activity steadily increases until midday 
and stay relatively consistent until decreasing once again around hour 20 and 
until hour 24 (likely when participants are getting ready to sleep). 
Interestingly, we see a slight steeper decline in activity beginning around 
hour 12 for participants with less than high school education compared to 
the other 2 categories. Generally, the trends in activity correspond to the 
awake hours of the participants. We see the activities reach the lowest 
in that hour 0-4 range, suggesting a time frame where the participants are 
consistently asleep.



## Problem 3

Clean January 2020 Dataset
```{r}
Jan_2020 <- 
  read_csv(
    file = "Data/citibike/Jan 2020 Citi.csv",
    na = c("NA", "", "."),
    col_select = c(ride_id:member_casual)) %>%
  janitor::clean_names() %>% 
  mutate(
    month_year = "January 2020"
  )

head(Jan_2020)

```

Clean January 2024 Dataset
```{r}
Jan_2024 <- 
  read_csv(
    file = "Data/citibike/Jan 2024 Citi.csv",
    na = c("NA", "", "."),
    col_select = c(ride_id:member_casual)) %>% 
  janitor::clean_names() %>% 
  mutate(
    month_year = "January 2024"
  )

head(Jan_2024)


```

Clean July 2020 Dataset
```{r}
July_2020 <- 
  read_csv(
    file = "Data/citibike/July 2020 Citi.csv",
    na = c("NA", "", "."),
    col_select = c(ride_id:member_casual)) %>% 
  janitor::clean_names() %>%
  mutate(
    month_year = "July 2020"
  )
  

head(July_2020)

```

Clean July 2024 Dataset
```{r}
July_2024 <- 
  read_csv(
    file = "Data/citibike/July 2024 Citi.csv",
    na = c("NA", "", "."),
    col_select = c(ride_id:member_casual)) %>% 
  janitor::clean_names() %>%
  mutate(
    month_year = "July 2024"
  )

head(July_2024)


```
Combine the datasets 
```{r}
Jan_July_20_24 <- rbind(Jan_2020, July_2020, Jan_2024, July_2024) %>% 
  mutate(
    weekdays = factor(
      weekdays,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", 
                 "Saturday", "Sunday")
    )
  )

head(Jan_July_20_24)

```



Create reader-friendly table showing total number of rides in each combination
of year and month separating casual riders and Citi Bike members
```{r}
Jan_July_20_24 %>% 
  select(month_year, member_casual) %>% 
  group_by(month_year, member_casual) %>% 
  count() %>% 
  pivot_wider(
    names_from = month_year,
    values_from = n
  ) %>%
  relocate("July 2020", .after = "January 2020") %>% 
  knitr::kable()



```
Looking at the table, we can see that usage of Citi bike seems to wane 
in January for both casual users and membership holders, irrespective of the 
year of the data. This makes sense as one would assume bike usage decreases in
the winter. It is also interesting to note that the number of uses of Citi bike
in the months of January and July increased from 2020 to 2024 for both the 
casual user and membership holder groups.

5 most popular starting stations for July 2024
```{r}
July_2024 %>% 
  select(start_station_name) %>% 
  group_by(start_station_name) %>% 
  count() %>% 
  arrange(desc(n)) %>% 
  head(5) %>% 
  rename(
    "number_uses" = n
  ) %>% 
  knitr::kable()


```

plot to investigate the effects of day of the week, month, and year on 
median ride duration
```{r}
Jan_July_20_24 <- Jan_July_20_24 %>% 
  separate(month_year, c("month", "year"), sep = " ")

ggplot(Jan_July_20_24, aes(x = weekdays, y = duration, fill = month)) +
  geom_violin(alpha = 0.5) +
  stat_summary(
    aes(group = month),
    fun = "median", 
    color = "blue", 
    position = position_dodge(width = 0.9),
    show.legend = FALSE) +
  xlab("Day of the Week") +
  ylab("Duration of Bike Ride") +
  facet_grid(.~year)



```
Looking at the series of violin plots. There are couple of interesting things 
to point out. Firstly, regardless of day of week, or month, or year, the 
violin plots for duration of bike ride demonstrate significant right skew. 
We can see that most values clusters between 0 and 25, yet a large amount 
of upper-bound outliers skew the distribution. This is why we see these long, 
thin tails moving upwards from each violin plot. Secondly, it is interesting 
to note that while the median duration of ride is greater in July for both 
2020 and 2024, from a visual scan, it seems that the gap has actually 
decreased from 2020 to 2024. Furthermore, we see that the portions of the 
violin plot with more substantial concentration of data points seem longer in 
2020 than 2024. I wonder if this relates to the increase in usage of Citi bikes
from 2020 to 2024. Perhaps there is a sampling effect at play, where 
in 2020, more avid bikers (either casual or membership-holding) were using Citi 
bike and thus had longer bike rides. By 2024, with a greater influx of riders,
perhaps these new riders dragged the median values downwards. Lastly,
I would like to note that there does not seem to be a large amount of variation
in the duration of bike ride by day (based solely on the median). The medians
are relatively similar for each day of the week, even when segmenting into 
the two differing months and the two different years.


Figure that shows the impact of month, membership status, and bike type on the 
distribution of ride duration
```{r}
Jan_July_20_24 %>% 
  filter(year == "2024") %>% 
ggplot(aes(x = month, y = duration, fill = member_casual )) +
  geom_violin(alpha = 0.5) +
  stat_summary(
    fun = "median", 
    color = "blue",
    position = position_dodge(0.9)) +
  xlab("Month of the Year") +
  ylab("Duration of Bike Ride") +
  facet_grid(.~rideable_type)

```
Looking at these two plots, we can first notice that based on the median 
duration of Bike Ride, casual users of Citi Biker seem to go on slightly longer
bike rides when using the classic bike compared to membership-holders. It must 
be stated that the difference is quite small. Regarding the electronic bike, 
based on the median duration of bike ride, there does not seem to be much 
difference between the casual user and the membership-holder. The results 
stated thus far are irrespective of the month of the year. Now if we examine 
the exact shapes of the violin plots, we can once again notice some differences.
Interestingly, the membership holder had bike-ride durations clustered closer 
to 0 for both the months of January and July. The shape of the violin plot 
for casual users of the classic bike in January and July are a bit different.
They are more evenly spread out, such that there is a more even concentration 
of values from 0 to ~30. This trend did not follow through with the violin plots 
for causal users' and membership-holders' usage of the electric bike during
January and July. In these cases, the shapes of the violin plots were very 
similar. Lastly, it appears that all these violin plots show a positive skew.
